The importance of a complete statistical description of dependence between variables for ecology and related fields

### 2019-05-02 Paper.Rmd
small edits

knitr::kable(tab_data_info, 
             format="latex", align="l",
             caption = "Summary table for the data we used. Bolds entries are the multivariate data, the rest are bivariate datasets (see Data section). Basal metabolic rate=BMR \\label{tab_data_info}", 
             booktabs = T,linesep = "\\addlinespace") %>%
             column_spec(1:4,width="2.8 cm")%>%
             row_spec(5:8, bold=T)
```

@@

To make results interpretable
to readers unfamiliar with copulas, we first introduce the normal copula and 
some other copulas we used; and we operationally introduce measures of tail dependence
and asymmetry of tail dependence. 
Conceptually, a bivariate copula is essentially a probability density function (pdf)
on the unit square with uniform maginals. 
Strictly speaking, 
the copula is actually defined as the bivariate 
cdf of the distribution rather than as the pdf,
but this makes little conceptual difference. 
The normal copula family is a one-parameter
family of copulas, with parameter corresponding to the correlation of the related 
bivariate normal distribtion, and controlling the degree of dependence 
(Fig. \ref{fig_cop_pedag3}A-B for an example).
Left- and right-tail dependence are $0$ for the normal copula, and this 
serves as a benchmark for tail dependence in other copulas - positive values of our
tail dependence measures correspond to dependence stronger than that of a normal
copula. The Clayton and 
survival-Clayton families (Fig. \ref{fig_cop_pedag1}G-H) are two more one-parameter 
families. The Clayton has positive lower-tail dependence ($0.87$ for
the example of Fig. \ref{fig_cop_pedag1}G) and $0$ upper-tail dependence; the 
survival-Clayton, which is a rotation of the Clayton by $180^\circ$, has 
positive upper-tail dependence ($0.87$ in Fig. \ref{fig_cop_pedag1}H)
and $0$ lower-tail dependence. For our model selection procedure (see below), we used 
several one- and two-parameter families of copulas exhibiting a wide variety of 
positive lower-tail dependence values 
(Clayton, survival-Gumbel, survival-Joe, survival-BB6, henceforth
abbreviated, respectively, C, SG, SJ, SBB6) and upper-tail dependence values
(survival-Clayton, Gumbel, Joe, BB6, henceforth abbreviated, respectively, 
SC, G, J, BB6). We also considered families that could exhibit positive upper- or lower-tail
dependence, or both, depending on parameters (BB1, survival-BB1, BB7, survival-BB7, henceforth 
abbreviated, respectively, BB1, SBB1, BB7, SBB7). 
If LT and UT are used to denote 
lower- and upper-tail dependence, their difference is a measure of asymmetry of 
tail dependence. We also used two families that were asymmetric in their non-tail parts, 
but that had zero tail dependence (BB8, survival-BB8, abbreviated BB8 and SBB8).
In addition to the normal copula, we used another symmetric copula,
the Frank copula (abbreviated F; Fig. \ref{fig_cop_pedag3}C-D). 
Copulas are implemented in the `copula` and `VineCopula` packages for the 
R statistical computing language.
Additional details are in introductory texts 
on copulas [@nelsen2006_copula ; @joe2014_dependence] and in `VineCopula` documentation.
Visualizations of the copulas we used are in Figs \ref{SM-figped_N}-\ref{SM-figped_SBB8}.

@@

We addressed Q1 principally via a model selection procedure in which the normal copula
the alternative copulas listed above were fitted to 
our ecological datasets. The resulting fits of the various copula models 
were compared via the Akaike and Bayesian 
Information Criteria (AIC and BIC). For bivariate datasets 
$(x_i,y_i)$ for $i=1,...,n$, model selection involved several steps. 
First, as in the Introduction, we produced ranks 
$(\tilde{u}_i,\tilde{v}_i)$ where $\tilde{u}_i$ is the rank of $x_i$ in the set 
$\{x_i:i=1,...,n\}$ and $\tilde{v}_i$ is the rank of $y_i$ in the set $\{y_i:i=1,...,n\}$.
Ranks were $1$ for the smallest element in a set. We then 
set $u_i=\tilde{u}_i/(n+1)$ and $v_i=\tilde{v}_i/(n+1)$; subsequent analysis used these 
*normalized ranks* $u_i$ and $v_i$. 
Second, we did a test of the independence of the $u_i$ and $v_i$ 
(implemented in `BiCopIndTest` in the `VineCopula` package). 
Subsequent model 
selection algorithms are ineffective if data could not be distinguished from independent data 
since most or all bivariate copula families we considered include the independent copula in the 
family or as a boundary case. If independence could be rejected ($0.05$ significance level), model 
selection proceeded. Third, we fit the bivariate copula families listed previously to the ranked data 
$(u_i,v_i)$ via maximum likelihood, using the `BiCopEst` function in `VineCopula`, and 
obtained AIC and BIC values and accompanying model weights, $\text{AIC}_{\text{w}}$ and 
$\text{BIC}_{\text{w}}$ [@burnham2003_modelselection]. The weight of a model represents the 
relative support of data for the model; weights sum to 1 across models considered. The lower- and 
upper-tail dependence of the best-fitting member of each family was also provided by `BiCopEst`. 
Fourth, $\text{AIC}_{\text{w}}$ values for all fitted models were used to get
model-averaged lower- and upper-tail dependence values using standard model averaging 
formulas [@burnham2003_modelselection]; likewise for BIC.
Model selection methods will give the *relative* support of several models, but will not,
on their own, indicate whether any of the models selected are an objectively good fit: all 
models considered might poorly describe the data. To test this, and as the final step of
our procedure, we tested the goodness of fit of our AIC-best copula family using a 
bootstrapping procedure implemented in `BiCopGofTest` of `VineCopula`. The method performed 
two tests, one based on a Cramer-von Mises statistic, and one based on a 
Kolmogorov-Smirnov statistic. To keep computation times reasonable, 
an initial run using $100$ bootstraps was performed, and if
the $p$-value resulting from either test was less than $0.2$, tests were re-run with 
$1000$ boostraps.

@@

Unlike $\cor_{l_b,u_b}$ and 
$\Ps_{l_b,u_b}$, for which large values inidicate strong dependence
in the portions of the the distributions of $u$ and $v$ given by 
the bounds, small values of 
$\Dsq_{l_b,u_b}$ indicate strong dependence. 
For large datasets (large $n$),
we could use $l_b$ and $u_b$ close together without incurring undue
sampling variation in our statistics, and we could consider 
multiple bands $(l_b,u_b)$ to understand how dependence
varies in different parts of the distributions. But for datasets with 
smaller $n$ we considered only $l_b=0$, $u_b=0.5$ and $l_b=0.5$, $u_b=1$.
We use the shorthand $\cor_l=\cor_{0,0.5}$ ($l$ is for "lower") 
and $\cor_u=\cor_{0.5,1}$ ($u$ is for "upper"); likewise $\Ps_l=\Ps_{0,.5}$,
$\Ps_u=\Ps_{0.5,1}$, $\Dsq_l=\Dsq_{0,0.5}$, and $\Dsq_u=\Dsq_{0.5,1}$.
To test for asymmetry of dependence in upper and lower portions of distributions,
we used differences of upper and lower statistics: for smaller datasets, 
$\cor_l-\cor_u$, $\Ps_l-\Ps_u$ and $\Dsq_u-\Dsq_l$ (note the opposite order
in the last of these); for large datasets, $\cor_{0,u_b}-\cor_{1-u_b,1}$, 
$\Ps_{0,u_b}-\Ps_{1-u_b,1}$ and
$\Dsq_{1-u_b,1}-\Dsq_{0,u_b}$ (again note the opposite order) with $u_b$ close 
to $0$.
We tested our statistics in Appendix \ref{SM-Test_npa_stat} (see also Figs \ref{SM-fig_stat_testing35}-\ref{SM-fig_stat_testing1000}).

@@

<!--Discuss stuff here on tradeoff and life history that Bever told me about-->
The relationships we studied between BMR and body mass relate
to a trade-off between mass-specific BMR (i.e., BMR per unit body mass) and body
mass itself. Trade-offs between life history characteristics are a focal topic in 
evolutionary ecology. There are reasons to believe copulas interrelate with 
life-history theory in additional ways beyond what we could demonstrate 
with the data we used.
It is well known that energy allocation to a life function, X (e.g., reproduction)
will reduce the energy that can be allocated to other functions, $Y_1$, $Y_2$, $Y_3$ 
(e.g., growth, predation avoidance; this is sometimes called the principle of 
allocation).
But it is equally well known that X can trade-off against any or all of 
the $Y_i$. Therefore, for large values of X which approach absolute 
limitations, there may be a strong 
association between X and $Y_1$. Whereas for small values of X there may be
little association - resources not allocated to X can instead be allocated 
to $Y_1$ or $Y_2$ or to some combination. This constitutes 
asymmetric tail dependence between X and $Y_1$. For instance, \cite{Winemiller1992}
described a three-way trade-off in North American fishes between age of reproductive
maturity, juvenile survivorship, and fecundity. The trade-off should
produce a tight association between age of maturity and fecundity for 
fishes with low age of maturity, but little association between those
variables for later-maturing
fishes because those fishes may invest much or little in juvenile survival.
Although copulas may be very useful for studying multi-dimensional trade-offs, 
applications will 
require careful attention to the possible consequences of biased sampling:
if the degree of completeness of a dataset is associated with one or
more of the quantities being studied then statistical artefacts
may bias conclusions of copula analyses (Appendix \ref{SM-sect:missingdata}, 
Fig. \ref{SM-fig:missingdata}).

>>>>

knitr::kable(tab_data_info, 
             format="latex", align="l",
             caption = "Summary table for the data we used. Bolds entries are the multivariate data, the rest are bivariate datasets (see Data section). Basal metabolic rate=BMR. \\label{tab_data_info}", 
             booktabs = T,linesep = "\\addlinespace") %>%
             column_spec(1:4,width="2.8 cm")%>%
             row_spec(5:8, bold=T)
```

@@

To make results interpretable
to readers unfamiliar with copulas, we first introduce the normal copula and 
some other copulas we used; and we operationally introduce measures of tail dependence
and asymmetry of tail dependence. 
Conceptually, a bivariate copula is a probability density function (pdf)
on the unit square with uniform maginals. 
The copula is actually defined as the bivariate 
cdf of the distribution rather than as the pdf itself,
but this makes little difference. 
The normal copula family is a one-parameter
family of copulas, with parameter corresponding to the correlation of the related 
bivariate normal distribtion, and controlling the degree of dependence 
(Fig. \ref{fig_cop_pedag3}A-B for an example).
Left- and right-tail dependence are $0$ for the normal copula, and this 
serves as a benchmark for tail dependence in other copulas - positive values of our
tail dependence measures correspond to dependence stronger than that of a normal
copula. The Clayton and 
survival-Clayton families (Fig. \ref{fig_cop_pedag1}G-H) are two more one-parameter 
families. The Clayton has positive lower-tail dependence ($0.87$ for
the example of Fig. \ref{fig_cop_pedag1}G) and $0$ upper-tail dependence; the 
survival-Clayton, which is a rotation of the Clayton by $180^\circ$, has 
positive upper-tail dependence ($0.87$ in Fig. \ref{fig_cop_pedag1}H)
and $0$ lower-tail dependence. For our model selection procedure (see below), we used 
several one- and two-parameter families of copulas exhibiting a wide variety of 
positive lower-tail dependence values 
(Clayton, survival-Gumbel, survival-Joe, survival-BB6, henceforth
abbreviated, respectively, C, SG, SJ, SBB6) and upper-tail dependence values
(survival-Clayton, Gumbel, Joe, BB6, henceforth abbreviated, respectively, 
SC, G, J, BB6). We also considered families that could exhibit positive upper- or lower-tail
dependence, or both, depending on parameters (BB1, survival-BB1, BB7, survival-BB7, henceforth 
abbreviated, respectively, BB1, SBB1, BB7, SBB7). 
If LT and UT are used to denote 
lower- and upper-tail dependence, their difference is a measure of asymmetry of 
tail dependence. We also used two families that were asymmetric in their non-tail parts, 
but that had zero tail dependence (BB8, survival-BB8, abbreviated BB8 and SBB8).
In addition to the normal copula, we used another symmetric copula,
the Frank copula (abbreviated F; Fig. \ref{fig_cop_pedag3}C-D). 
Copulas are implemented in the `copula` and `VineCopula` packages for the 
R statistical computing language.
Additional details are in introductory texts 
on copulas [@nelsen2006_copula ; @joe2014_dependence] and in `VineCopula` documentation.
Visualizations of the copulas we used are in Figs \ref{SM-figped_N}-\ref{SM-figped_SBB8}.

@@

We addressed Q1 principally via a model selection procedure in which the normal copula
and the alternative copulas listed above were fitted to 
our ecological datasets. The resulting fits of the various copula models 
were compared via the Akaike and Bayesian 
Information Criteria (AIC and BIC). For bivariate datasets 
$(x_i,y_i)$ for $i=1,...,n$, model selection involved several steps. 
First, as in the Introduction, we produced ranks 
$(\tilde{u}_i,\tilde{v}_i)$ where $\tilde{u}_i$ is the rank of $x_i$ in the set 
$\{x_i:i=1,...,n\}$ and $\tilde{v}_i$ is the rank of $y_i$ in the set $\{y_i:i=1,...,n\}$.
Ranks were $1$ for the smallest element in a set. We then 
set $u_i=\tilde{u}_i/(n+1)$ and $v_i=\tilde{v}_i/(n+1)$; subsequent analysis used these 
*normalized ranks* $u_i$ and $v_i$. 
Second, we did a test of the independence of the $u_i$ and $v_i$ 
(implemented in `BiCopIndTest` in the `VineCopula` package). 
Subsequent model 
selection algorithms are ineffective if data could not be distinguished from independent data 
since most or all bivariate copula families we considered include the independent copula in the 
family or as a boundary case. If independence could be rejected ($0.05$ significance level), model 
selection proceeded. Third, we fit the bivariate copula families listed previously to the ranked data 
$(u_i,v_i)$ via maximum likelihood, using the `BiCopEst` function in `VineCopula`, and 
obtained AIC and BIC values and accompanying model weights, $\text{AIC}_{\text{w}}$ and 
$\text{BIC}_{\text{w}}$ [@burnham2003_modelselection]. The weight of a model represents the 
relative support of data for the model; weights sum to 1 across models considered. The lower- and 
upper-tail dependence of the best-fitting member of each family was also provided by `BiCopEst`. 
Fourth, $\text{AIC}_{\text{w}}$ values for all fitted models were used to get
model-averaged lower- and upper-tail dependence values using standard model averaging 
formulas [@burnham2003_modelselection]; likewise for BIC.
Model selection methods will give the *relative* support of several models, but will not,
on their own, indicate whether any of the models selected are an objectively good fit: all 
models considered might poorly describe the data. To test this, and as the final step of
our procedure, we tested the goodness of fit of our AIC-best copula family using a 
bootstrapping procedure implemented in `BiCopGofTest` of `VineCopula`. The method performed 
two tests, one based on a Cramer-von Mises statistic, and one based on a 
Kolmogorov-Smirnov statistic. To keep computation times reasonable, 
an initial run using $100$ bootstraps was performed, and if
the $p$-value resulting from either test was less than $0.2$, tests were re-run with 
$1000$ boostraps.

@@

Unlike $\cor_{l_b,u_b}$ and 
$\Ps_{l_b,u_b}$, for which large values inidicate strong dependence
in the portions of the the distributions of $u$ and $v$ given by 
the bounds, small values of 
$\Dsq_{l_b,u_b}$ indicate strong dependence. 
For large datasets (large $n$),
we could use $l_b$ and $u_b$ close together without incurring undue
sampling variation in our statistics, and we could consider 
multiple bands $(l_b,u_b)$ to understand how dependence
varies in different parts of the distributions. But for datasets with 
smaller $n$ we considered only $l_b=0$, $u_b=0.5$ and $l_b=0.5$, $u_b=1$.
We use the shorthand $\cor_l=\cor_{0,0.5}$ ($l$ is for "lower") 
and $\cor_u=\cor_{0.5,1}$ ($u$ is for "upper"); likewise $\Ps_l=\Ps_{0,.5}$,
$\Ps_u=\Ps_{0.5,1}$, $\Dsq_l=\Dsq_{0,0.5}$, and $\Dsq_u=\Dsq_{0.5,1}$.
To test for asymmetry of dependence in upper and lower portions of distributions,
we used differences of upper and lower statistics: for smaller datasets, 
$\cor_l-\cor_u$, $\Ps_l-\Ps_u$ and $\Dsq_u-\Dsq_l$ (note the opposite order
in the last of these); for large datasets, $\cor_{0,u_b}-\cor_{1-u_b,1}$, 
$\Ps_{0,u_b}-\Ps_{1-u_b,1}$ and
$\Dsq_{1-u_b,1}-\Dsq_{0,u_b}$ (again note the opposite order) with $u_b$ close 
to $0$.
We tested our statistics in Appendix \ref{SM-Test_npa_stat} (see also Figs \ref{SM-fig_stat_testing35}-\ref{SM-fig_stat_testing1000}). All code for this
project is archived at \url{www.github.com/sghosh89/BIVAN}.

@@

<!--Discuss stuff here on tradeoff and life history that Bever told me about-->
The relationships we studied between BMR and body mass relate
to a trade-off between mass-specific BMR (i.e., BMR per unit body mass) and body
mass itself. Trade-offs between life history characteristics are a focal topic in 
evolutionary ecology. There are reasons to believe copulas interrelate with 
life-history theory in additional ways beyond what we could demonstrate 
with the data we used.
It is well known that energy allocation to a life function, X (e.g., reproduction)
will reduce the energy that can be allocated to other functions, $Y_1$, $Y_2$, $Y_3$ 
(e.g., growth, predation avoidance; this is sometimes called the principle of 
allocation).
But it is equally well known that X can trade-off against any or all of 
the $Y_i$. Therefore, for large values of X which approach absolute 
limitations, there may be a strong 
association between X and $Y_1$. Whereas for small values of X there may be
little association - resources not allocated to X can instead be allocated 
to $Y_1$ or $Y_2$ or to some combination. This constitutes 
asymmetric tail dependence between X and $Y_1$. For instance, \cite{Winemiller1992}
described a three-way trade-off in North American fishes between age of reproductive
maturity, juvenile survivorship, and fecundity. The trade-off should, in theory,
produce a tight association between age of maturity and fecundity for 
fishes with low age of maturity, but little association between those
variables for later-maturing
fishes because those fishes may invest much or little in juvenile survival.
Although copulas may be very useful for studying multi-dimensional trade-offs, 
applications will 
require careful attention to the possible consequences of biased sampling:
if the degree of completeness of a dataset is associated with one or
more of the quantities being studied then statistical artefacts
may bias conclusions of copula analyses (Appendix \ref{SM-sect:missingdata}, 
Fig. \ref{SM-fig:missingdata}).
